---
title: "Concordance"
author: "Sarah Power"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: pdf_document
---

# Recipient
Heather Finkle.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
mfrow=c(1, 2) 
```
```{r data entry}
# load ----
source('distribution_functions.R')
library(lubridate)

# data ----
read_csv('../data/ChigISGrunappt2006-2017catch.by.district.csv') %>% 
  dplyr::select(-Earlycatch, -Latecatch, -Earlytotal, -Latetotal) %>%
  dplyr::rename(#date = Date,
         prop_early_genetics = Propotionearly, 
         early_esc_genetics = Earlyesc, 
         late_esc_genetics = Lateesc,
         chignik = Chignik_Hook_Kujulik) %>%
  mutate(esc = early_esc_genetics + late_esc_genetics,
         #catch = early_catch_genetics + late_catch_genetics,
         #catch_chignik = Lagoon + chignik,
         #run = esc + catch_chignik, #catch is our estimate of what would have occured at the wier had there been no fishing, based on migration timing studies.
         #run_early_gen = prop_early_genetics*run,
         date_weir = lubridate::mdy(Date),
         day_of_year = lubridate::yday(date_weir)) -> chig_data # convert the Date to its numeric equivalent

harvest_data <- read_csv('../data/harvest19692017.csv') %>% 
  filter(year %in% c(2006:2008)) %>%
  mutate(date_harvest = mdy(date_begin),
         day_of_year_harvest = yday(date_harvest),
         weir_arrival = weir_date(area, date_harvest), 
         allocation_percent = allocation(area, date_harvest)
         )

head(harvest_data)

catch <-aggregate( fish_count ~ weir_arrival + area, data=harvest_data, FUN=sum)
area <- unique(catch$area)
area <- c("Central", "Chignik", "Dolgoi", "Eastern", "Igvak", "Ikatan", "Perryville", "SEDM", "Shumagins" )
allocation <- c()
as_data_frame()

year_vector <- c(2006:2008,2010:2017)
```

# Background



This analysis examines run timing distributions and how well they can distinguish early run stocks from later stocks on run timing using various ways of counting harvest. The results will be compared to the run timing distributions that are determined with the additional aid of genetics to examine how appropriate it is to use certain harvests.

Currently the run consists of the escapement as captured by the weir, and catches from various management districts, some of which are distant from Chignik river.  

# Method

Creelman (2011) hypothesized that the early run and later runs are comprised of various smaller runs with their own run timing and run locations. She found that run timing played an important role in genetic variation. This makes distinguishing run based on run timing a possibility. She grouped collections into four genetic groups: Black Lake, Chignik Lake, Chignik Lake October, and Chignik River.An initial review of run timing across years indicates that there are typically three district run timing modes, not four.(Appendix A, original distributions to be fit.) Noting that data from the weir is collected from about late May until the end of September, and that Creelman noted an October Chignik Lake run the assumption is that there is a biological reason for three modes to be presented in the data.

Maximum likelihood methods including the EM algorithm and a Newton-type algorithm to parse out various distributions using the mixdist package in R. 

Distributions of the early run were estimated using run timing and genetic assignment for each year. This run timing estimate was used as the baseline for comparison of the models evaluated sans genetics.

In initial investigations Normal and Weibull distributions were fit to the multimodal run timing data, additionally certain constraints were considered in order to find the best fit. Those constraints included equal standard errors, means equally spaced, and having the mean or standard error of the early distribution fixed. The fixed means and standard errors were the medians values from the early distribution determined using genetics.

The most appropriate model was a trimodal, normally distributed model with the constraint of equal standard deviations for each of the distributions. This model had an early run timing with means and standard deviations most like the estimated early run timing distributions determined with genetics.

Once this model was determined to be most appropriate the run counts of each of the years were examined in order to determine if the model could be useful. For 8 of the 11 years between 2006 and 2017 early run counts using run timing alone were within 10% of the counts using genetic information. The three best years included 2010, 2013, and 2014, where early run counts using run timing information alone differed by no more than 3% of the counts using additional genetic information. (Shown below) 


```{r three_best, fig.width = 6, fig.height = 4}
auto_year(weir_data, 2010)
year_stats(weir_data, 2010)
auto_year(weir_data, 2013)
year_stats(weir_data, 2013)
auto_year(weir_data, 2014)
year_stats(weir_data, 2014)

```

\pagebreak 
The three years that run timing distributions were off by more than 10% included 2008, which overestimated the early run by 27%, 2012 which overestimated by 67% and 2015 which underestimated by 25% and 2012.


```{r three_worst, fig.width = 6, fig.height = 4}
auto_year(weir_data, 2008)
year_stats(weir_data, 2008)
auto_year(weir_data, 2012)
year_stats(weir_data, 2012)
auto_year(weir_data, 2015)
year_stats(weir_data, 2015)

```
\pagebreak

It was suggested by Heather Finkle that using catch information from farther flung management districts might reduce the ability to accurate model run timing since those districts would contain harvest from non-Chignik stocks. The analysis was reexamined this time using only harvests that were within a two day fish travel time from Chignik weir. This change improved the model significantly for 2012. The run timing only model previously over estimated thegenetics model by 67% after the management area correction this dropped to a 5% over estimate. The 2008 and 2015 estimates were similar to those found under the original model with a 23% overestimate and 25% underestimate respectively.

\pagebreak

```{r even_better, fig.width = 6, fig.height = 4}
auto_year(chig_data, 2012)
year_stats(chig_data, 2012)
```

# Results

For most years the model 

# Recommendations/Discussion

Future analysis could include 

# Citation

Elisabeth K. Creelman, Lorenz Hauser, Ryan K. Simmons, William D. Templin & Lisa W. Seeb (2011):Temporal and Geographic Genetic Divergence: Characterizing Sockeye Salmon Populations in the Chignik Watershed, Alaska,
Using Single-Nucleotide Polymorphisms, Transactions of the American Fisheries Society, 140:3, 749-762

Link to this article: http://dx.doi.org/10.1080/00028487.2011.584494

\pagebreak

# R Session Info
```{r session_info}
sessionInfo()
```

\pagebreak

# Appendix A
Below are the graphs for all the years using model with the more conservative data set:

```{r all_year_graphs, fig.width = 6, fig.height = 4}
auto_year(chig_data, 2006)
year_stats(chig_data, 2006)
auto_year(chig_data, 2007)
year_stats(chig_data, 2007)
auto_year(chig_data, 2008)
year_stats(chig_data, 2008)
auto_year(chig_data, 2010)
year_stats(chig_data, 2010)
auto_year(chig_data, 2011)
year_stats(chig_data, 2011)
auto_year(chig_data, 2012)
year_stats(chig_data, 2012)
auto_year(chig_data, 2013)
year_stats(chig_data, 2013)
auto_year(chig_data, 2014)
year_stats(chig_data, 2014)
auto_year(chig_data, 2015)
year_stats(chig_data, 2015)
auto_year(chig_data, 2016)
year_stats(chig_data, 2016)
auto_year(chig_data, 2017)
year_stats(chig_data, 2017)

```